Relationships in the extremes and their influence on competition and coexistence

### 2021-07-29 paper/Paper.Rnw
comments
added captions

\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

%***DAN: stopped here 2021 07 20
[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature. 
Interacting environmental drivers can be asymmetrically related in the extreme values (as shown in \cite{Gosh_2020_AER}). 
Environmental drivers directly influence competitive pressure experienced by populations through environmental response; competition is always a function of environmental response \citep{Chesson_2012, Ellner_2016}.
Thus we have implemented our tools for the study of ATAs to explore the importance of environmental relationships in their extremes on competition and coexistence. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, hence we continue/move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. Moreover, the lottery model within the framework of coexistence theory provides a suitable opportunity to explore relationships in their extremes and the consequences to coexistence.

To flesh out the idea of how ATAs in environmental response might manifest and modfiy the outcome of coexistence, we introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}. The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by the resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and unfavorable environment and can persist in the community. When rain conditions fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. This produces a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated. ATAs in environmental response can be caused by extreme climatic events. If the community is affected by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Conversely, if the community is effected by frequent and long floods with some years of normalcy, the grass-forb environmental relationship may become more correlated in the higher values or right-tail associated. In Figure 2 we see a possible outcome of ATAs between environmental responses. In simulating their populations,  ATAs can cause species to have longer periods of extreme dominance and lower growth rates overall.

@@

\begin{figure}
\caption{Figure displaying the noise. Three panels, left-, symmetric, and right-tail assoc noise. Use standard normal marginals.
Explain in caption that alternatives for $(b_1,b_2)$ are obtained by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.
Jasmin to make and insert the figure here using the includegraphics command, and then draft a caption along the lines described here.\label{fig:noise}}
\end{figure}

\begin{figure}
\caption{Pedagogical figure somehow displaying the different outcomes of competition in ATA and symmetric cases for an example.
This will be cited from the Intro, so needs to be understandable to someone who does not know any of the ideas we introduce yet.
So in particular you cannot use IGR as your metric of coexistence, you need to use something more intuitive. You will pick
model parameter values and do two sims, one with ATA noise and one with symmtric noise, and show different outcomes. We should
discuss and possibly take several iterations to get this figure right since it's a pedagogically important figure - if the reader
is not convinced in the Intro that we are studying something that makes a difference, they are not going to slog through
the Theory section, which is inevitably going to be tough going for many readers.\label{fig:pedagog}}
\end{figure}
%possible examples:
%population simulations - shows symmetric with faster fluctatuations to extremes of coexistence while asymmetric has longer periods of noticeable coexistence and dominace/exclusion; may be unclear

>>>>

\begin{document}
\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}

@@

%***DAN: stopped here 2021 07 20
[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature.
%;we review these studies here 
Interacting environmental drivers can be asymmetrically related in the extreme values (as shown in \cite{Gosh_2020_AER}). 
Environmental drivers directly influence competitive pressure experienced by populations through environmental response; competition is always a function of environmental response \citep{Chesson_2012, Ellner_2016}.
Thus we have implemented our tools for the study of ATAs to explore the importance of environmental relationships in their extremes on competition and coexistence. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, hence we continue/move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. Moreover, the lottery model within the framework of coexistence theory provides a suitable opportunity to explore relationships in their extremes and the consequences to coexistence.

If extreme events are more temporally extensive/expansive than regular environmental fluctuations, extreme events may cause the environmental response of two species who depend on temporally separated indicators of growth to become more correlated in the extremes. For example, ...

To demonstrate plausibility of ATA altering coexistence between species, we carried out long time scale simulations using a simple lottery model. 
To flesh out the idea of how ATAs in environmental response might manifest and modfiy the outcome of coexistence, we introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}. The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by the resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and unfavorable environment and can persist in the community. When rain conditions fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. This produces a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated. ATAs in environmental response can be caused by extreme climatic events. If the community is affected by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Conversely, if the community is effected by frequent and long floods with some years of normalcy, the grass-forb environmental relationship may become more correlated in the higher values or right-tail associated. In Figure 2 we see a possible outcome of ATAs between environmental responses. In simulating their populations,  ATAs can cause species to have longer periods of extreme dominance and lower growth rates overall.

@@

\begin{figure}
\caption{
%Figure displaying the noise. Three panels, left-, symmetric, and right-tail assoc noise. Use standard normal marginals.Explain in caption that alternatives for $(b_1,b_2)$ are obtained by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.Jasmin to make and insert the figure here using the includegraphics command, and then draft a caption along the lines described here.
Simple example of bivariate noise with standard normal marginals. A) is left-tail associated, B) is symmetric, and C) is right-tail associated. All three cases have the same Pearson correlation of $\rho=$, up to sampling.., yet show different symmetry of tail associaton. Asymmetrically tail associated cases are perfectly correlated in the lower (A)  or upper (C) halfs of the distribution. The variables $(b_1,b_2)$ are transformed by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.
\label{fig:noise}}
\end{figure}

\begin{figure}
\caption{%Pedagogical figure somehow displaying the different outcomes of competition in ATA and symmetric cases for an example. This will be cited from the Intro, so needs to be understandable to someone who does not know any of the ideas we introduce yet. So in particular you cannot use IGR as your metric of coexistence, you need to use something more intuitive. You will pick model parameter values and do two sims, one with ATA noise and one with symmtric noise, and show different outcomes. We should discuss and possibly take several iterations to get this figure right since it's a pedagogically important figure - if the reader is not convinced in the Intro that we are studying something that makes a difference, they are not going to slog through the Theory section, which is inevitably going to be tough going for many readers.
Populations inlfuenced by asymmetric tail association...
recover from low densities less often/ spend more time on average under one-species dominance. A) is a simulation over time of a population within a two species lottery model whose environmental response and competition is influenced by symmetric bivariate noise. B) is like A), but the population is influenced by bivariate noise with left-tailed asymmetric association. Both simulations were ran for total time of 100,000. Noticeable population density was defined at a threshold of 0.95 of total community space. Periods of noticable coexistence are marked in grey. While the population in B) had longer mean periods of coexistence, it had much longer mean periods of one-species dominace. The population in B) also recovered from low densities less often and thus has a lower ability to recover when rare and thus that community has lower coexistence.     
\label{fig:pedagog}}
\end{figure}
